Monte Carlo as Approximate Polynomial Preconditioning within MCSA

### 2014-05-13 doc/polynomial_note/polynomial.tex
Adding discussion of computing GMRES polynomial coefficients.

\documentclass[10pt]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[labelfont=bf,labelsep=period]{caption}
\usepackage{booktabs}
\usepackage{multirow}

@@

\newcommand{\bx}{\ensuremath{\mathbf{x}}}
\newcommand{\bb}{\ensuremath{\mathbf{b}}}
\newcommand{\br}{\ensuremath{\mathbf{r}}}
\newcommand{\bp}{\ensuremath{\mathbf{p}}}
\newcommand{\bw}{\ensuremath{\mathbf{w}}}
\newcommand{\bA}{\ensuremath{\mathbf{A}}}
\newcommand{\bH}{\ensuremath{\mathbf{H}}}
\newcommand{\bM}{\ensuremath{\mathbf{M}}}
\newcommand{\bP}{\ensuremath{\mathbf{P}}}
\newcommand{\bW}{\ensuremath{\mathbf{W}}}
\newcommand{\bI}{\ensuremath{\mathbf{I}}}

@@

A significant downside to the use of Chebyshev polynomials is the need for
accurate bounds on the spectrum of $\bA$.  For nonsymmetric matrices,
this requires not only computing smallest and largest eigenvalues as is
required in the symmetric case but requires
bounding the entire spectrum of $\bA$ in an ellipse.  The process for
computing such bounding ellipses is described in Ref.~\cite{manteuffel_77}.
An alternative class of polynomials requiring less spectral information
while handling nonsymmetric systems naturally is the polynomials generated
by GMRES \cite{saad_86}.  The GMRES algorithm computes, at iteration $m$,
an approximate solution to \eqref{eq:lin_problem}, $\bx^{(m)}$, 
such that the $L_2$ norm of the residual
is minimized over the $(m+1)$-dimensional Krylov subspace given by
\begin{equation}
K_{m+1}(\bA,\bb) = \text{span}\{ \bb, \bA \bb, \ldots, \bA^{m} \bb \}.
\end{equation}
From this definition it is clear that GMRES is constructing (implicitly)
a polynomial of degree $m$, $p_m(\bA)$, such that $p_m(\bA) \approx \bA^{-1}$.
Use of the the GMRES polynomial as a preconditioner to fixed point iterations
was studied in Refs.~\cite{nachtigal_92,joubert_94} among several others
and as a preconditioner to {GMRES} itself in Ref.~\cite{liu_14}.

@@

As a simple test problem, we choose the ``JPWH 991'' matrix from
the Harwell-Boeing Collection, available from the Matrix Market.  
The sparsity pattern of the JPWH matrix is given in Figure~\ref{fig:jpwh_sparsity}.
\begin{figure}
\includegraphics[width=4.5in]{jpwh_sparsity}
\caption{Sparsity pattern of JPWH matrix \label{fig:jpwh_sparsity}}
\end{figure}
The matrix is nonsymmetric, but its eigenvalues are all real and
contained in the interval $[-16.3,-0.12]$.  After a diagonal row
scaling, the eigenvalues are contained in the interval $[0.02,1.71]$.
All test cases use the diagonally scaled matrix and the right
hand side of the linear system is taken to be a vector of all ones.
The stopping criterion is a reduction in the L$_2$ norm of the residual
by six orders of magnitude.

@@

\section{Questions}

\begin{enumerate}

\item Using $\bI - \bA$ as a basis for the polynomial expansion rather
than $\bA$ results in a much more stable approach with either
Neumann series or Chebyshev polynomials.  What property of $\bH$
(or the resulting $\bP$ and $\bW$ matrices) is responsible for this
and is there a basis that is superior to $\bH$?

\item Chebyshev polynomials offer significant reduction in iteration
count for a given polynomial order relative to the Neumann series polynomial, 
but require significantly more
histories for stability.  Is there another class of polynomials that
might improve on the Neumann series without requiring as many
histories as Chebyshev?
\end{enumerate}

>>>>

\documentclass[10pt]{article}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage[margin=1in]{geometry}
\usepackage[labelfont=bf,labelsep=period]{caption}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{subcaption}

@@

\newcommand{\bx}{\ensuremath{\mathbf{x}}}
\newcommand{\by}{\ensuremath{\mathbf{y}}}
\newcommand{\bb}{\ensuremath{\mathbf{b}}}
\newcommand{\bc}{\ensuremath{\mathbf{c}}}
\newcommand{\br}{\ensuremath{\mathbf{r}}}
\newcommand{\bp}{\ensuremath{\mathbf{p}}}
\newcommand{\bw}{\ensuremath{\mathbf{w}}}
\newcommand{\bA}{\ensuremath{\mathbf{A}}}
\newcommand{\bH}{\ensuremath{\mathbf{H}}}
\newcommand{\bM}{\ensuremath{\mathbf{M}}}
\newcommand{\bP}{\ensuremath{\mathbf{P}}}
\newcommand{\bW}{\ensuremath{\mathbf{W}}}
\newcommand{\bI}{\ensuremath{\mathbf{I}}}
\newcommand{\bK}{\ensuremath{\mathbf{K}}}
\newcommand{\calK}{\ensuremath{\mathcal{K}}}

\DeclareMathOperator*{\argmin}{arg\,min}

@@

A significant downside to the use of Chebyshev polynomials is the need for
accurate bounds on the spectrum of $\bA$.  For nonsymmetric matrices,
this requires not only computing smallest and largest eigenvalues as is
required in the symmetric case but requires
bounding the entire spectrum of $\bA$ in an ellipse.  The process for
computing such bounding ellipses is described in Ref.~\cite{manteuffel_77}.
An alternative class of polynomials requiring less spectral information
while handling nonsymmetric systems naturally is the polynomials generated
by GMRES \cite{saad_86}.  The GMRES algorithm computes, at iteration $m$,
an approximate solution to \eqref{eq:lin_problem}, $\bx^{(m)}$,
such that the $L_2$ norm of the residual
is minimized over the $(m+1)$-dimensional Krylov subspace given by
\begin{equation}
\calK_{m+1}(\bA,\bb) = \text{span}\{ \bb, \bA \bb, \ldots, \bA^{m} \bb \}.
\end{equation}
From this definition it is clear that GMRES is constructing (implicitly)
a polynomial of degree $m$, $p_m(\bA)$, such that $p_m(\bA) \approx \bA^{-1}$.
Use of the the GMRES polynomial as a preconditioner to fixed point iterations
(so-called hybrid GMRES methods)
was studied in Refs.~\cite{nachtigal_92,joubert_94} among several others
and as a preconditioner to {GMRES} itself in Ref.~\cite{liu_14}.

In the context of hybrid GMRES methods, the coefficients of the GMRES polynomial
are generally not computed explicitly, rather the polynomial is applied
to a vector using relationships between the Krylov basis vectors arising from
the Arnoldi process.  In order to use a polynomial within the Monte Carlo
process, however, an explicit representation of the polynomial coefficients
is required.  Multiple approaches for computation of these coefficients
are possible.  One possibility is to observe that the eigenvalues of the
upper Hessenberg matrix generated by the Arnoldi process are the roots of the
GMRES polynomial.  Therefore the polynomial can be written as
\begin{equation}
p_m(x) = \prod_{i=1}^{m} (x - \lambda_i) \:, \label{eq:prod_gmres_poly}
\end{equation}
where $\lambda_i$ are the eigenvalues of the Hessenberg matrix.
Computing the coefficients in the form of Eq.~\eqref{eq:power_poly}
is then simply a matter of expanding the terms in Eq.~\eqref{eq:prod_gmres_poly}.
A second method for computing the GMRES polynomial coefficients arises
directly from the GMRES minimization problem:
\begin{align}
\bx^{(m)} = \argmin_x \Vert \bb - \bA \bx \Vert_2 \\
\text{s.t.} \quad \bx \in \calK_{m+1}(\bA,\bb) \:.
\end{align}
By choosing the ``na\"ive'' basis for the Krylov subspace,
\begin{equation}
\bK_{m+1} = \left[ \bb, \bA \bb, \ldots, \bA^m \bb \right] \:,
\end{equation}
the coeffcients of the GMRES polynomial are given by the solution to
the least squares problem
\begin{equation}
\min_{\bc} \Vert \bA \bK_{m+1} \bc - \bb \Vert_2 \:. \label{eq:gmres_min}
\end{equation}
In Ref.~\cite{liu_14}, the surprising observation was made that solving
Eq.~\eqref{eq:gmres_min} using the normal equations actually produced
a more accurate result than using a QR factorization to solve
Eq.~\eqref{eq:gmres_min} or using Eq.~\eqref{eq:prod_gmres_poly}.
In our experience, the selection of the preferred method is less clear
cut.  Figure~\ref{fig:gmres_poly} shows the GMRES polynomial with coefficients
computed using either Eq.~\eqref{eq:prod_gmres_poly} or by
Eq.~\eqref{eq:gmres_min} with two different least squares solvers.
The matrix \bA is a $25 \times 25$ 1-D Laplacian with $0.05$ added
to each diagonal entry and the initial vector is taken to be a
normally-distributed random vector.  At a polynomial order of 5,
the two least squares methods are equivalent and appear to perform
significantly better than the Hessenberg eigenvalue approach
(i.e. the magnitude of the polynomials are smaller over the majority
of the region of interest).  At a polynomial order of 10, the behavior
is similar to the $m=5$ case except that a slight difference between the
two least square solutions is now evident.  All methods perform fairly
well for $m=15$, although the normal equations approach is noticably
larger over portions of the domain.  Finally, for $m=20$ the
eigenvalue-based approach appears the best with the least squares QR
now appearing to suffer.  Table~\ref{tab:gmres_poly} shows the value
of the spectral radius of $\bI - p_m(\bA) \bA$ for each
of these cases.  Although the eigenvalue-based approach
appears to perform worse at small polynomial orders, it actually
produces the smallest spectral radius for all four cases and is the
only approach to produce a spectral radius that is monotonically decreasing
with polynomial order.
The dramatic variability in the different methods for computing the
coefficients of the GMRES polynomial are indicative of the ill-conditioned
nature of computations involving polynomial coefficients.
\begin{figure}
\centering
\begin{subfigure}[b]{0.49\linewidth}
\includegraphics[width=\textwidth]{gmres_poly5}
\caption{$m = 5$}
\end{subfigure}
\begin{subfigure}[b]{0.49\linewidth}
\includegraphics[width=\textwidth]{gmres_poly10}
\caption{$m = 10$}
\end{subfigure}
\begin{subfigure}[b]{0.49\linewidth}
\includegraphics[width=\textwidth]{gmres_poly15}
\caption{$m = 15$}
\end{subfigure}
\begin{subfigure}[b]{0.49\linewidth}
\includegraphics[width=\textwidth]{gmres_poly20}
\caption{$m = 20$}
\end{subfigure}
\caption{Plot of $q_{m+1}(x)=1-xp_{m}(x)$, where $p_m(x)$ is the GMRES
 polynomial with different methods for computing the coefficients.
 \bA\ is a shifted Laplacian with $N=25$.  Markers
 indicate the polynomials evaluated at the eigenvalues of \bA.}
\label{fig:gmres_poly}
\end{figure}

\begin{table}
\centering
\caption{Spectral Radius of $\left( \bI - p_m(\bA) \bA \right)$.}
\label{tab:gmres_poly}
\begin{tabular}{ccccc}
\toprule
& \multicolumn{4}{c}{Polynomial Order} \\
\cmidrule(r){2-5}
Method & 5 & 10 & 15 & 20 \\
\midrule
Hessenberg Eigenvalues     & 0.569 & 0.231 & 0.109 & 0.088 \\
Least Squares Normal Eqns. & 0.632 & 0.259 & 0.184 & 0.247 \\
Least Squares QR           & 0.632 & 0.273 & 0.176 & 0.586 \\
\bottomrule
\end{tabular}
\end{table}

@@

As a simple test problem, we choose the ``JPWH 991'' matrix from
the Harwell-Boeing Collection, available from the Matrix Market.
The sparsity pattern of the JPWH matrix is given in Figure~\ref{fig:jpwh_sparsity}.
\begin{figure}
\centering
\includegraphics[width=4.5in]{jpwh_sparsity}
\caption{Sparsity pattern of JPWH matrix \label{fig:jpwh_sparsity}}
\end{figure}
The matrix is nonsymmetric, but its eigenvalues are all real and
contained in the interval $[-16.3,-0.12]$.  After a diagonal row
scaling, the eigenvalues are contained in the interval $[0.02,1.71]$.
All test cases use the diagonally scaled matrix and the right
hand side of the linear system is taken to be a vector of all ones.
The stopping criterion is a reduction in the L$_2$ norm of the residual
by six orders of magnitude.

@@
